---
title: 'Data Analysis Assignment #1'
author: "Pasaye, Joshua"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

***##### Section 1: Summarizing the data.***

(1)(a) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
# Use summary() function to obtain and present descriptive statistics
summary(mydata)

# Use table() function for CLASS and RINGS
table(mydata$CLASS)

table(mydata$RINGS)

```

**Question:  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: Variables SEX and CLASS are both categorical or qualitative variables. All other variables are numeric or quantitative variables. HEIGHT appears to be the variable closest to a symmetrical normal distribution, with mean and median fairly similar and a skewness of -0.23. LENGTH (skewness=-.67) and DIAM (skewness=-0.62) are both negatively skewed, which suggest the presence of low outliers causing the mean to be lower than the median. WHOLE (skewness=.47), SHUCK (skewness=0.64), RINGS (skewness=1.24), VOLUME (skewness=.44), and RATIO (skewness=.72) are all positively skewed, which suggests the presence of high outliers causing the mean to be greater than the median.***

(1)(b) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
# Create contingency table of sex & class
SEX <- mydata$SEX
CLASS <- mydata$CLASS
contingency_table <- addmargins(table(SEX, CLASS))
contingency_table

# Create barplot of the data
ggplot(data = mydata, aes(x = CLASS, fill = SEX)) + # Add data and mapping, Class on x axis and color by SEX
  geom_bar(stat = "count", position = "dodge", color = "black") + # Add bar chart, unstack bars, make black borders
  labs(title = "Class Membership, SEX-differentiated", # Change title
       y = "Frequency") + # Change y axis label
  scale_y_continuous(breaks = seq(0, 140, by = 20)) + # Change y axis scale
  scale_fill_manual(values = c("darkred", "darkblue", "darkgreen"), # Change legend value colors
                    labels = c("Female", "Infant", "Male")) + # Change legend value labels
  theme(legend.title = element_blank(), # Remove legend title
        legend.box.background = element_rect(), # Add legend box
        legend.background = element_blank(), # Remove legend background
        panel.background = element_blank(), # Remove plot panel background
        legend.position = "inside", # Move legend inside plot
        legend.position.inside = c(.85, .90), # Move legend box
        panel.grid.major = element_blank(), # Remove major grid lines
        panel.border = element_blank(), # Remove panel border
        axis.ticks.x = element_blank(), # Remove x axis
        axis.line.y = element_line(linewidth = .5, linetype = "solid") # Change y axis line size
        )
```

**Essay Question:  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: The male and female groups follow a normal distribution, though there may be some negative skewness in the A4 and A5 class. The infant group is positively skewed with most of the data around the A1 & A2 class. Lastly, there seem to be more males than females in each class.***

(1)(c) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
# set seed
set.seed(123)

# Simple random sample of mydata
work <- mydata[sample(x = nrow(mydata), size = 200), ] # select rows from mydata then select those sampled rows
work

# Create scatter plot of work
plot(work[, 2:6])
```

-----

***##### Section 2: Summarizing the data using graphics.***

(2)(a) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
ggplot(data = mydata, aes(x = VOLUME, y = WHOLE)) + # Add data and mapping, VOLUME on x, WHOLE on y
  geom_point(aes(color = CLASS)) + # Create scatter plot, distinguish by CLASS
  labs(title = bquote('Whole weight, in grams | Volume, in'~ cm^3), # Change title
       x = 'Volume', # Change x axis label
       y = 'Whole weight') + # Change y axis label
  scale_color_brewer(name = "BuPu") + # Add colors
   theme(legend.title = element_blank(), # Remove legend title
        legend.box.background = element_rect(), # Create legend box
        legend.background = element_blank(), # Remove legend background
        legend.position = "inside", # Change legend position to inside plot
        legend.position.inside = c(.85, .30), # Move legend box
        panel.background = element_blank(), # Remove panel background
        axis.ticks = element_line(linewidth = .75, linetype = 1), # Change axis tick size
        axis.line = element_line(linewidth = .5, linetype = 1) # Change axis line size
        )
```


(2)(b) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
ggplot(data = mydata, aes(x = WHOLE, y = SHUCK)) + # Add data and mapping, WHOLE on x, SHUCK on y
  geom_point(aes(color = CLASS)) + # Create scatter plot, distinguish by CLASS
  geom_abline(linetype = 2, intercept = 0, slope = max(mydata$SHUCK/mydata$WHOLE)) + # Add line at 0 intercept, and max ratio for slope
  labs(title = 'Shuck weight, in grams | Whole weight, in grams', # Change title
       x = 'Whole weight', # Change x axis label
       y = 'Shuck weight') + # Change y axis label
  scale_color_manual(values = c("lightgoldenrod", "tan", "red", "darkorange3", "chocolate4")) + # Add colors
   theme(legend.title = element_blank(), # Remove legend title
        legend.box.background = element_rect(), # Create legend box
        legend.background = element_blank(), # Remove legend background
        legend.position = "inside", # Change legend position to inside plot
        legend.position.inside = c(.80, .25), # Move legend box
        panel.background = element_blank(), # Remove panel background
        axis.ticks = element_line(linewidth = .75, linetype = 1), # Change axis tick size
        axis.line = element_line(linewidth = .5, linetype = 1) # Change axis line size
        )

```

**Essay Question:  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: When a comparison is made between the two graphs, they are very similar in trend: the higher the value of the x-axis then so too does the y-axis. The first graph plots volume against whole weight, cm^3 against g. The older classes, A4 and A5, have their starting volumes at 200cm^3. For most classes that are younger, their volume tends to be smaller. This shows that the younger abalones are smaller in volume thus smaller in whole weight. The older abalones are larger in volume thus larger in whole weight. Now, shuck weight is being compared to whole weight; g vs. g. Older classes A4 and A5 show larger whole weights with a smaller shuck weight corresponding to each. Most of the younger classes show larger shuck weights with smaller whole weights. This, in fact, tells us that even though the younger abalones weigh less than the older ones, you get more meat (shucked weight) out of them compared to the older abalones.***

-----

***### Section 3: Getting insights about the data using graphs.***

(3)(a) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a, fig.height= 7, fig.width= 7}
# Create graph
par(mfrow = c(3,3)) # Create grid
hist(subset(mydata$RATIO, mydata$SEX == "F"), main = "Female RATIO", col = "red", xlab = "") # Create female histogram
hist(subset(mydata$RATIO, mydata$SEX == "I"), main = "Infant RATIO", col = "green", xlab = "") # Create infant histogram
hist(subset(mydata$RATIO, mydata$SEX == "M"), main = "Male RATIO", col = "blue", xlab = "") # Create male histogram
boxplot(subset(mydata$RATIO, mydata$SEX == "F"), col = "red") # Create female boxplot
title(main = "Female RATIO") # Add title
boxplot(subset(mydata$RATIO, mydata$SEX == "I"), col = "green") # Create infant boxplot
title(main = "Infant RATIO") # Add title
boxplot(subset(mydata$RATIO, mydata$SEX == "M"), col = "blue") # Create male boxplot
title(main = "Male RATIO") # Add title
qqnorm(subset(mydata$RATIO, mydata$SEX == "F"), main = "Female RATIO", col = "red") # Create qq plot
qqline(subset(mydata$RATIO, mydata$SEX == "F")) # Add normal line
qqnorm(subset(mydata$RATIO, mydata$SEX == "I"), main = "Infant RATIO", col = "green") # Create qq plot
qqline(subset(mydata$RATIO, mydata$SEX == "I")) # Add normal line
qqnorm(subset(mydata$RATIO, mydata$SEX == "M"), main = "Male RATIO", col = "blue") # Create qq plot
qqline(subset(mydata$RATIO, mydata$SEX == "M")) # Add normal line
```

**Essay Question: Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: From the above histograms, female and infant are generally normally distributed. Male, however seems to have a result that is a little-skewed with a spike in the data around 0.135 gm/cm^3. From the boxplots it can be seen that there exist some outliers. Male and infant appear to each have some outliers in the upper ratio values. But female has outliers both in the upper and lower ends of the data. This is also confirmed in the qqplots. In each qqplot, there are several points at the far right of the qqline that don't fall on the qqline. It is also obvious on the far left of the female qqplot. Also, in the qqplots for one male and one infant, the data is deviating away from the qqline far to the right, which may indicate some skewness.***

(3)(b) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
# Infant classification
i_out <- mydata[mydata$SEX == "I" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "I"])$out, ]

# Female classification
f_out <- mydata[mydata$SEX == "F" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "F"])$out, ]

# Male classification
m_out <- mydata[mydata$SEX == "M" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "M"])$out, ]

# Combining use rbind and kable
kable(rbind(i_out, f_out, m_out))

```

**Essay Question:  What are your observations regarding the results in (3)(b)?**

***Answer: Most of the shuck/volume ratio outliers come from the infant classification. Further, the large majority of outliers are from the younger classes, A1 - A3; only one outlier is classified as A5. In addition, the only extreme outliers come from the female and infant classes, but none appear for the male class. The number of outliers in these younger classes might be skewing the data and make drawing correlation between physical measurements and age difficult.***

-----

***### Section 4: Getting insights about possible predictors.***

(4)(a) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
# Create plots
b_v <- ggplot(data = mydata, aes(x = CLASS, y = VOLUME)) +
  geom_boxplot() # Boxplots of VOLUME
b_w <- ggplot(data = mydata, aes(x = CLASS, y = WHOLE)) +
  geom_boxplot() # Boxplots of WHOLE
s_v <- ggplot(data = mydata, aes(x = RINGS, y = VOLUME)) +
  geom_point() # Scatter plot of VOLUME
s_w <- ggplot(data = mydata, aes(x = RINGS, y = WHOLE)) +
  geom_point() # Scatter plot of WHOLE

grid.arrange(b_v, b_w, s_v, s_w, nrow = 2)
```

**Essay Question: How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: Neither volume nor whole weight is a particularly good predictor of age in abalone. Both are positively correlated with the number of rings, but the correlation is weak-for example, an abalone of 500 cm^3 volume may have from 8 to 23 rings. Similarly, classifications tend to be older with increasing volume and weight, but the interquartile ranges of classes A3, A4, and A5 overlap considerably, making it hard to determine the age. This would suggest either that volume and weight are not good estimators of age or something is wrong with the classification system, but again emphasizes that there are complexities with identifying the correct ring count.***

-----

***### Section 5: Getting insights regarding different groups in the data.***

(5)(a) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
# Aggreate data
agg <- aggregate(mydata[c("VOLUME", "SHUCK", "RATIO")], by = list(SEX, CLASS), FUN = mean)

# Create matrix of VOLUME data
v <- matrix(data = agg$VOLUME, nrow = 3, ncol = 5, dimnames = list(c("F", "I", "M"), 
                                                                   list("A1", "A2", "A3", "A4", "A5"))
            )

# Create matrix of SHUCK data
s <- matrix(data = agg$SHUCK, nrow = 3, ncol = 5, dimnames = list(c("F", "I", "M"),
                                                                  list("A1", "A2", "A3", "A4", "A5"))
            )

# Create matrix of RATIO data
r <- matrix(data = agg$RATIO, nrow = 3, ncol = 5, dimnames = list(c("F", "I", "M"),
                                                                  list("A1", "A2", "A3", "A4", "A5"))
            )

# Use kable to present matrices
kable(v, caption = "Volume")
kable(s, caption = "Shuck")
kable(r, caption = "Ratio")
```

(5)(b) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.height = 15, fig.width = 6}
# Create RATIO vs CLASS plot
r_c <- ggplot(data = agg, aes(x = Group.2, y = RATIO, group=Group.1, color=Group.1)) + # Use agg dataset with mean, put Group.2 (CLASS) on x axis, RATIO on y axis, group and color by Group.1 (SEX)
  geom_line(linewidth= 1) + # Add line plot, change line width
  geom_point(size = 3) + # Add points to plot, change size of points
  labs(title = "Mean RATIO per CLASS", # Change title 
       x = "CLASS") + # Change x axis label
  scale_color_manual(name = "Sex", # Change legend title
                     values = c("red", "blue", "green"), # Change color values
                     breaks = c("F", "I", "M"), # Indicates SEX variables
                     labels = c("Female", "Infant", "Male") # Change legend labels
                     )

# Create VOLUME vs CLASS plot
v_c <- ggplot(data = agg, aes(x = Group.2, y = VOLUME, group=Group.1, color=Group.1)) + # Use agg dataset with mean, put Group.2 (CLASS) on x axis, VOLUME on y axis, group and color by Group.1 (SEX)
  geom_line(linewidth=1) + # Add line plot, change line width
  geom_point(size = 3) +  # Add points to plot, change size of points
  labs(title = "Mean VOLUME per CLASS", # Change title
       x = "CLASS") + # Change x axis label
  scale_color_manual(name = "Sex", # Change legend title
                     values = c("red", "blue", "green"), # Change color values
                     breaks = c("F", "I", "M"), # Indicates SEX variables
                     labels = c("Female", "Infant", "Male") # Change legend labels
                     )

# Create SHUCK vs CLASS plot
s_c <- ggplot(data = agg, aes(x = Group.2, y = SHUCK, group=Group.1, color=Group.1)) + # Use agg dataset with mean, put Group.2 (CLASS) on x axis, SHUCK on y axis, group and color by Group.1 (SEX)
  geom_line(linewidth=1) + # Add line plot, change line width
  geom_point(size = 3) + # Add points to plot, change size of points
  labs(title = "Mean SHUCK per CLASS", # Change title
       x = "CLASS") + # Change x axis label
  scale_color_manual(name = "Sex", # Change legend title
                     values = c("red", "blue", "green"), # Change color values
                     breaks = c("F", "I", "M"), # Indicates SEX variables
                     labels = c("Female", "Infant", "Male") # Change legend labels
                     )

# Use grid.arrange to combine plots
grid.arrange(r_c, v_c, s_c, nrow = 3, ncol=1)
```

**Essay Question:  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: In general, across all sexes, we see that with increasing class the average shuck/volume ratio decreases, and both volume and shuck increase with class. However, these plots beg a number of questions regarding the differences in the way that shuck and volume vary between genders. Namely, why do females tend to be heavier and larger than both males and infants? Further, why do females tend to be heavier/larger but their growth over time is smaller than males and infants?***

5(c) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
# Create infant volume by rings boxplot
v_r_i <- ggplot(data = subset(mydata, mydata$RINGS < 16 &  mydata$SEX == "I"), aes(x = factor(RINGS), y = VOLUME)) + # Subset data for infant sex and rings below 16, put rings on x axis and volume on y axis
  geom_boxplot(fill = "blue", color = "black") + # Create boxplot, add blue color and black border
  coord_cartesian(ylim = c(0, 1100)) + # Change ylim from 0 to 1100
  labs(title = "Infant Volume | Rings", # Change title
       x = "Rings", # Change x axis label
       y = "Volume") # Change y axis label

# Create adult volume by rings boxplot
v_r_a <- ggplot(data = subset(mydata, mydata$RINGS < 16 &  mydata$SEX != "I"), aes(x = factor(RINGS), y= VOLUME)) + # Subset data for male or female sex and rings below 16, put rings on x axis and volume on y axis
  geom_boxplot(fill = "red", color = "black") + # Create boxplot, add red color and black border
  coord_cartesian(ylim = c(0, 1100)) + # Change ylim from 0 to 1100
  labs(title = "Adult Volume | Rings", # Change title
       x = "Rings", # Change x axis lablel
       y = "Volume") # Change y axis label

# Create infant whole weight by rings boxplot
w_r_i <- ggplot(data = subset(mydata, mydata$RINGS < 16 &  mydata$SEX == "I"), aes(x = factor(RINGS), y = WHOLE)) + # Subset data for infant sex and rings below 16, put rings on x axis and whole weight on y axis
  geom_boxplot(fill = "blue", color = "black") + # Create boxplot, add blue color and black borders
  coord_cartesian(ylim = c(0, 400)) + # Change ylim from 0 to 400
  labs(title = "Infant Whole weight | Rings", # Change title
       x = "Rings", # Change x axis label
       y = "Whole weight") # Change y axis label

# Create adult whole weight by rings boxplot
w_r_a <- ggplot(data = subset(mydata, mydata$RINGS < 16 &  mydata$SEX != "I"), aes(x = factor(RINGS), y = WHOLE)) + # Subset data for male or female sex and rings below 16, put rings on x axis and whole weight on y axis
  geom_boxplot(fill = "red", color = "black") + # Create boxplot, add red color and black borders
  coord_cartesian(ylim = c(0, 400)) + # Change ylim from 0 to 400
  labs(title = "Adult Whole weight | Rings", # Change title
       x = "Rings", # Change x axis label
       y = "Whole weight") # Change y axis label

# Combine plots onto a single page
grid.arrange(v_r_i, v_r_a, w_r_i, w_r_a, nrow = 2)
```

**Essay Question:  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: These graphs show that between about 1 ring and 10-11 rings, abalones experience the greatest growth and as such volume and weight increase rapidly. After 10-11 rings, the rate of growth stabilizes or may even begin to decline. Typically speaking, infants do weigh and measure smaller than the adults; however, there does seem to be some overlap within the size and weight of the specimens, with some infants nearly the size of smaller adults. Besides, infants are less variable in volume and weight. It can be seen that the interquartile range and the standard deviation are smaller than the more variable adults.***


-----

***### Section 6: Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: The original study likely concluded in failure due to the misclassification of immature abalone. It is extremely challenging to discern between immature and small adult abalone. Accurate sex information is a key determinant in predicting age from physical measurements since females weigh more than males in younger abalone, in general. Growth in size and weight is reduced after 10-11 rings, making it tough to estimate the accurate age. Thus, wrong sex classification reduces the accuracy of age prediction in younger abalone, and growth slowdowns in A3-A5-year-old abalone further obscures age determination based on physical metrics.***


**Essay Question 2) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: I would also like to inquire about information regarding the sample and population: whether the sampling was random or non-random, how the sampling frame was developed, when the sample was drawn-data relevance, sample size in relation to the population size, known population parameters, and possible biases that may affect representativeness. Several factors may determine whether a sample is truly representative of the population.***

**Essay Question 3)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: The data analysis of observational studies is rather complex, since there are a great many influential factors. Errors and biases in human judgment in designing the study, in data collection alone, or even in data analysis alone often produce greatly differing results. Such studies often cannot control for several important explanatory variables, and lacking a control group, any relationship found cannot establish cause and effect. Still, such studies can suggest variables of interest that may be further studied for cause.***
