---
title: "Data Analysis Assignment #2"
author: "Pasaye, Joshua"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


## Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(dplyr)
library(flux)
library(ggplot2)
library(gridExtra)
library(knitr)
library(rockchalk)
library(tidyverse)

# 2) Use the "mydata.csv" file from Assignment #1 
# Reading the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file structure.

mydata <- read.csv("mydata.csv", sep = ",", stringsAsFactors = TRUE)
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```


***#### Section 1: ####***

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a, fig.width = 12}
# Plot hist and QQ plot using RATIO data
par(mfrow = c(1,2))
hist(x = mydata$RATIO, breaks = 30, main = "Histogram of RATIO data", ylab = "Frequency", xlab = "Ratio", col = "darkred")
qqnorm(y = mydata$RATIO, main = "Normal RATIO Q-Q Plot", xlab = "Theoretical Quantiles",
        ylab = "Sample Quantiles", col = "darkred")
qqline(y = mydata$RATIO, col = "darkred")
par(mfrow = c(1,1))

# Calculate kurtosis and skewness
sprintf("The skewness of RATIO is %.4f", rockchalk::skewness(mydata$RATIO))
sprintf("The kurtosis of RATIO is: %.4f", rockchalk::kurtosis(mydata$RATIO))

```

(1)(b) Transform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b, fig.width = 12, fig.height = 8}
# Transform RATIO data
mydata$L_RATIO <- log10(mydata$RATIO)

# Plot histogram and qq plot
layout(matrix(c(1, 2,  # First, second
                3, 3), # and third plot
            nrow = 2,
            ncol = 2,
            byrow = TRUE)
       )

hist(x = mydata$L_RATIO, breaks = 30, main = "Histogram of Log10 RATIO data", xlab = "Log10 RATIO", 
     ylab = "Frequency", col = "darkred")
qqnorm(y = mydata$L_RATIO, main = "Normal Log10 RATIO Q-Q Plot", xlab = "Theoretical Quantiles",
       ylab = "Sample Quantiles", col = "darkred")
qqline(y = mydata$L_RATIO, col = "darkred")
boxplot(mydata$L_RATIO ~ mydata$CLASS, main = "Boxplot of Log10 RATIO data by CLASS", 
        xlab = "CLASS", ylab = "Log10 RATIO", col = c("pink2", "darkorange", "darkgreen", "blue3", "purple3"))
par(mfcol = c(1,1))

# Calculate skewness and kurtosis
sprintf("The skewness of Log10 RATIO is %.4f", rockchalk::skewness(mydata$L_RATIO))
sprintf("The kurtosis of Log10 RATIO is: %.4f", rockchalk::kurtosis(mydata$L_RATIO))
```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
# Perform bartlett test of homogeneity
bartlett.test(L_RATIO ~ CLASS, data = mydata)

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: Based on the three previous steps, the L_RATIO variable exhibits better conformance to a normal distribution because the skewness for L_RATIO was -.0939, which is relatively low skewness, whereas the skewness for RATIO was .715, which is a higher, more moderate amount of skewness. Similarly, the kurtosis for L_RATIO (~3.535) was lower than the RATIO kurtosis (~4.167), indicating that the transformed RATIO data has heavy tails and conforms more to a normal distribution. Visually this is depicted by the histogram and Q-Q plots between the two variables, the RATIO data has noticeable potential outliers at the tail end of the data making the data more right skewed, whereas, the L_RATIO histogram and Q-Q plots are more aligned to a normal distribution.***


***#### Section 2 ####***

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}
# Perform first ANOVA
summary(aov(L_RATIO ~ CLASS*SEX, data= mydata))

# Perform second ANOVA
summary(aov(L_RATIO ~ CLASS + SEX, data = mydata))
```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: The non-significant interaction between CLASS and SEX indicates that combined, they are not appropriate predictors of the model. Based on the p-values of the individual CLASS and SEX factors, they may be better predictors of the model when taken individually.***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}
# Perform HSD test on second model
TukeyHSD(aov(L_RATIO ~ CLASS + SEX, data = mydata))

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: The only coefficient that does not seem to be significant in the age classes is between A2 and A1. Besides those, all of them are significant. For Male vs female, we can reject the null hypothesis that male and female have no significant difference. This allows us to combine them in the Adult group.***


***####  Section 3: ####***

(3)(a1) Here, we will combine "M" and "F" into a new level, "ADULT". The code for doing this is given to you. For (3)(a1), all you need to do is execute the code as given.

```{r Part_3a1}
# Here, we show how to define the new variable TYPE using only base R functions:

mydata$TYPE <- factor(ifelse(mydata$SEX == "I", "I", "ADULT"))
table(mydata$TYPE)


```

(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2, fig.width = 12}
# Plot graphs
par(mfrow = c(1,2))
hist(mydata$VOLUME[mydata$TYPE == "I"], breaks = 30, main = "Infant Volume", xlab = "",
     ylab = "Volume", col = "darkgreen")
hist(mydata$VOLUME[mydata$TYPE == "ADULT"], breaks = 30, main = "Adult Volume", xlab = "",
     ylab = "Volume", col = "darkgreen")
par(mfrow = c(1,1))

```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: The distribution of Infants seems to be skewed to the right, while the distribution of adults seems to be normally distributed. The histograms show that the majority of the Infant’s volume seems to be from 0 to roughly 200, while the majority of the Adult’s volume appears to be from 300 to 600. It does not seem that there will be any difficulty separating adults from infants.***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b, fig.width = 12, fig.height = 8}
# Create plots
p1 <- ggplot(data = mydata, aes(x = VOLUME, y = SHUCK, color = CLASS)) +
  geom_point() +
  labs(fill = "") +
  theme(legend.position = "bottom")
p2 <- ggplot(data = mydata, aes(x = log10(VOLUME), y = log10(SHUCK), color = CLASS)) +
  geom_point() +
  labs(x = "L_VOLUME",
       y = "L_SHUCK",
       fill = "") +
  theme(legend.position = "bottom")
p3 <- ggplot(data = mydata, aes(x = VOLUME, y = SHUCK, color = TYPE)) +
  geom_point() +
  labs(fill = "") +
  theme(legend.position = "bottom")
p4 <- ggplot(data = mydata, aes(x = log10(VOLUME), y = log10(SHUCK), color = TYPE)) +
  geom_point() +
  labs(x = "L_VOLUME",
       y = "L_SHUCK",
       fill = "") +
  theme(legend.position = "bottom")
grid.arrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: Comparing the scatterplots, the log-transformation significantly reduces the variability in the data, especially at the higher values of the x-axis, resulting in a more even spread of points across the plot, which is particularly noticeable in the upper range of the y-axis. By stabilizing the variance, log-transformation can help meet the assumption of homoscedasticity in linear regression, leading to more reliable model estimations and hypothesis tests. In both plots, the different CLASS levels appear as distinct clusters along the x-axis, with lower CLASS levels generally located on the left side and higher CLASS levels on the right. The TYPE levels seem to be distributed throughout the entire range of the x-axis in both plots, with no clear visual separation between the different TYPE categories based solely on their position on the x-axis.***


***####   Section 4: ####***

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. You are given code in (4)(a1) to reclassify the infants in classes A4 and A5 as ADULTS. 

```{r Part_4a1}

mydata$TYPE[mydata$CLASS == "A4" | mydata$CLASS == "A5"] <- "ADULT"
table(mydata$TYPE)

```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}
# Create new variables with transformed SHUCK & VOLUME
mydata$L_SHUCK <- log10(mydata$SHUCK)
mydata$L_VOLUME <- log10(mydata$VOLUME)

# Perform Regression
my_model <- lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data = mydata)
summary(my_model)
```

**Essay Question:  Interpret the trend in CLASS level coefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: The CLASS levels coefficient estimates have a trend that is positively related to the outcome variable with increasing CLASS levels. In other words, as the CLASS level increases, the predicted value of the outcome variable typically increases in a relatively linear fashion. This is in agreement with the previous graphs, which indicated a positive relationship between the levels of CLASS and the outcome studied. This would imply that higher levels of CLASS are normally associated with better outcomes.***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: TYPE is fairly significant with an estimate of -.021, however, CLASS 4A and 5A have a higher estimate than TYPE, indicating that these CLASSES would be better predictors of L_SHUCK for decision-making.***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

***#### Section 5: ####***

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5am, fig.width = 12}
# Graph histogram and qq plot
par(mfrow = c(1, 2))
hist(x = my_model$residuals, breaks = 30, main = "", xlab = "Residuals", ylab = "Count",
     col = "royalblue")
qqnorm(y = my_model$residuals, main = "", xlab = "Theortical", ylab = "Sample",
       col = "royalblue")
qqline(y = my_model$residuals)
par(mfrow = c(1,1))

# Calculate skewness and kurtosis
sprintf("The skewness of the model residuals is %.4f", rockchalk::skewness(my_model$residuals))
sprintf("The kurtosis of the model residuals  is: %.4f", rockchalk::kurtosis(my_model$residuals))
```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b, fig.width = 12, fig.height = 8}
# Create plots
p5 <- ggplot(data = mydata, aes(x = L_VOLUME, y = my_model$residuals, color = CLASS)) +
  geom_point() +
  ylab("Residuals") +
  theme(legend.position = "bottom")
p6 <- ggplot(data = mydata, aes(x = L_VOLUME, y = my_model$residuals, color = TYPE)) +
  geom_point() +
  ylab("Residuals") +
  theme(legend.position = "bottom")
p7 <- ggplot(data = mydata, aes(x = CLASS, y = my_model$residuals)) +
  geom_boxplot(col = c("red3", "yellow3", "green3", "blue3", "purple3")) +
  ylab("Residuals")
p8 <- ggplot(data = mydata, aes(x = TYPE, y = my_model$residuals)) +
  geom_boxplot(col = c("darkred", "darkblue")) +
  ylab("Residuals")
grid.arrange(p5, p6, p7, p8, nrow = 2, ncol = 2)

# Test homogeneity
bartlett.test(x = my_model$residuals, g = mydata$CLASS)
```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: Based on the displays and calculations in (5)(a) and (5)(b), the model seems to fit reasonably well, indicating that "L_VOLUME" (and thus "VOLUME") could be a useful factor for making harvesting decisions. The scatterplot reveals a clear positive linear relationship between the predicted and actual values, while the residual plot shows a fairly even distribution around zero, suggesting that the model effectively captures the main patterns in the data without notable bias.***

-----

Harvest Strategy:

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.The Management needs to make a decision to implement 1 rule that meets the business goal.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

***#### Section 6: ####***

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

# Checking values
length(volume.value)
length(prop.infants)
length(prop.adults)
head(volume.value)
head(prop.infants)
head(prop.adults)
```

(6)(b) Our first "rule" will be protection of all infants. We want to find a volume cutoff that protects all infants, but gives us the largest possible harvest of adults. We can achieve this by using the volume of the largest infant as our cutoff. You are given code below to identify the largest infant VOLUME and to return the proportion of adults harvested by using this cutoff. You will need to modify this latter code to return the proportion of infants harvested using this cutoff. Remember that we will harvest any individual with VOLUME greater than our cutoff.

```{r Part_6b}
# Largest infant volume
(max_inf_vol <- max(mydata$VOLUME[mydata$TYPE == "I"]))  # [1] 526.6383

# Proportion of adults harvested
(max_adu_prop <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_inf_vol) /
  total.adults)  # [1] 0.2476573

# Add code to calculate the proportion of infants harvested
(max_inf_prop <- sum(mydata$VOLUME[mydata$TYPE == "I"] > max_inf_vol) /
  total.infants)

# If we use the largest infant volume, we harvest approximately 24.8% of adults and 0%,
# as expected, of infants.

```

(6)(c) Our next approaches will look at what happens when we use the median infant and adult harvest VOLUMEs. Using the median VOLUMEs as our cutoffs will give us (roughly) 50% harvests. We need to identify the median volumes and calculate the resulting infant and adult harvest proportions for both.

```{r Part_6c}
# Add code to determine the median infant volume:
(med_inf_vol <- median(mydata$VOLUME[mydata$TYPE == "I"]))

# Add code to calculate the proportion of infants harvested
(prop_inf_har1 <- sum(mydata$VOLUME[mydata$TYPE == "I"] > med_inf_vol) /
  total.infants)

# Add code to calculate the proportion of adults harvested
(prop_adu_har1 <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_inf_vol) /
  total.adults)

# If we use the median infant volume as our cutoff, we harvest almost 50% of our infants
# and a little more than 93% of our adults.


# Add code to determine the median adult volume:
(med_adu_vol <- median(mydata$VOLUME[mydata$TYPE == "ADULT"]))

# Add code to calculate the proportion of infants harvested
(prop_inf_har2 <- sum(mydata$VOLUME[mydata$TYPE == "I"] > med_adu_vol) /
  total.infants)

# Add code to calculate the proportion of adults harvested
(prop_adu_har2 <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_adu_vol)/
  total.adults)

# If we use the median adult volume as our cutoff, we harvest almost 50% of adults
# and approximately 2.4% of infants.

```

(6)(d) Next, we will create a plot showing the infant conserved proportions (i.e. "not harvested," the prop.infants vector) and the adult conserved proportions (i.e. prop.adults) as functions of volume.value. We will add vertical A-B lines and text annotations for the three (3) "rules" considered, thus far:  "protect all infants," "median infant" and "median adult." Your plot will have two (2) curves - one (1) representing infant and one (1) representing adult proportions as functions of volume.value - and three (3) A-B lines representing the cutoffs determined in (6)(b) and (6)(c).

```{r Part_6d, fig.width = 12, fig.height = 6}
# Create plot
ggplot() +
  geom_line(mapping = aes(x = volume.value, y = prop.infants, col = "Infants")) +
  geom_line(mapping = aes(x = volume.value, y = prop.adults, col = "Adults")) +
  labs(title = "Proportion of Adults and Infants Protected",
       x = "Volume",
       y = "Proportion",
       col = "") + 
  scale_color_manual(breaks = c("Adults", "Infants"),
                     values = c("darkblue", "darkred")) +
  geom_vline(xintercept = med_inf_vol) +
  annotate("text", label=paste("Median Infant\n", round(med_inf_vol, 1)), x=med_inf_vol+50, y=.45) +
  geom_vline(xintercept = med_adu_vol) +
  annotate("text", label=paste("Median Adult\n", round(med_adu_vol, 1)), x=med_adu_vol+50, y=.45) +
  geom_vline(xintercept = max_inf_vol) +
  annotate("text", label=paste("Protect all infants\n", round(max_inf_vol, 1)), x=max_inf_vol+60, y=.95) +
  geom_hline(yintercept = .50, linetype = 2) +
  theme(legend.position = "inside",
        legend.justification = c(.95, .87),
        legend.background = element_blank(),
        plot.title = element_text(face = "bold", size = 12, hjust = 0)
        )
```

**Essay Question:  The two 50% "median" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: There is a clear trade-off between harvesting adults and the risk of also harvesting infants. If we focus on protecting all infants, adult harvests will be minimal. Opting for the median adult allows for a higher adult harvest but also increases the risk of harvesting more infants. On the other hand, selecting the median number of infants will protect more of them but result in fewer adults being harvested. Therefore, a more balanced metric is needed to optimize adult harvesting while minimizing the harvesting of infants.***

-----


More harvest strategies:

This part will address the determination of a cutoff volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. In other words, we want to find the volume value such that the vertical distance between the infant curve and the adult curve is maximum. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

Note on ROC:

There are multiple packages that have been developed to create ROC curves. However, these packages - and the functions they define - expect to see predicted and observed classification vectors. Then, from those predictions, those functions calculate the true positive rates (TPR) and false positive rates (FPR) and other classification performance metrics. Worthwhile and you will certainly encounter them if you work in R on classification problems.
However, in this case, we already have vectors with the TPRs and FPRs. Our adult harvest proportion vector, (1 - prop.adults), is our TPR. This is the proportion, at each possible 'rule,' at each hypothetical harvest threshold (i.e. element of volume.value), of individuals we will correctly identify as adults and harvest. Our FPR is the infant harvest proportion vector, (1 - prop.infants). We can think of TPR as the Confidence level (ie 1 - Probability of Type I error and FPR as the Probability of Type II error. At each possible harvest threshold, what is the proportion of infants we will mistakenly harvest?
Our ROC curve, then, is created by plotting (1 - prop.adults) as a function of (1 - prop.infants). In short, how much more 'right' we can be (moving upward on the y-axis), if we're willing to be increasingly wrong; i.e. harvest some proportion of infants (moving right on the x-axis)?


-----

***#### Section 7:  ####***

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}
# Create difference
difference <- (1 - prop.adults) - (1 - prop.infants)
head(difference)

# Plot difference
ggplot() + 
  geom_line(aes(volume.value, difference), col = "darkblue") +
  labs(title = "Difference in Proportions Harvested",
  x = "Volume",
  y = "Difference in Proportions Harvested")

```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

# Add smooth line to plot
ggplot() + 
  geom_line(aes(volume.value, difference), col = "darkblue") +
  geom_smooth(aes(volume.value, smooth.difference), col = "darkred", linetype = 2) +
  labs(title = "Difference in Proportions Harvested",
  x = "Volume",
  y = "Difference in Proportions Harvested")

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

Include, side-by-side, the plot from (6)(d) but with a fourth vertical A-B line added. That line should intercept the x-axis at the "max difference" volume determined from the smoothed curve here.

```{r Part_7c, fig.width = 12, fig.height = 6}
# Present side-by-side plot
p9 <- ggplot() + 
  geom_line(aes(volume.value, difference), col = "darkblue") +
  geom_smooth(aes(volume.value, smooth.difference), col = "darkred", linetype = 2) +
  geom_vline(xintercept = volume.value[which.max(smooth.difference)]) +
  annotate("text", label=paste("Volume = ", round(volume.value[which.max(smooth.difference)], 3)),
           x=volume.value[which.max(smooth.difference)]+15, y=.4, angle = 90) +
  labs(title = "Difference in Proportions Harvested",
  x = "Volume",
  y = "Difference in Proportions Harvested")

p10 <- ggplot() +
  geom_line(mapping = aes(x = volume.value, y = prop.infants, col = "Infants")) +
  geom_line(mapping = aes(x = volume.value, y = prop.adults, col = "Adults")) +
  labs(title = "Proportion of Adults and Infants Protected",
       x = "Volume",
       y = "Proportion",
       col = "") + 
  scale_color_manual(breaks = c("Adults", "Infants"),
                     values = c("darkblue", "darkred")) +
  geom_vline(xintercept = med_inf_vol) +
  annotate("text", label=paste("Med Inf\n", round(med_inf_vol, 1)), x=med_inf_vol+60, y=.45) +
  geom_vline(xintercept = med_adu_vol) +
  annotate("text", label=paste("Med Adl\n", round(med_adu_vol, 1)), x=med_adu_vol+60, y=.45) +
  geom_vline(xintercept = max_inf_vol) +
  annotate("text", label=paste("Protect\n all inf\n", round(max_inf_vol, 1)), x=max_inf_vol+60, y=.94) +
  geom_vline(xintercept =  volume.value[which.max(smooth.difference)]) +
  annotate("text", label=paste("Max dif\n", round(volume.value[which.max(smooth.difference)], 1)),
             x = volume.value[which.max(smooth.difference)] +60, y = .80) +
  geom_hline(yintercept = .50, linetype = 2) +
  theme(legend.position = "inside",
        legend.justification = c(.95, .87),
        legend.background = element_blank(),
        plot.title = element_text(face = "bold", size = 12, hjust = 0)
        )
grid.arrange(p9, p10, nrow = 1)
```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions. We will actually calculate these proportions in two ways:  first, by 'indexing' and returning the appropriate element of the (1 - prop.adults) and (1 - prop.infants) vectors, and second, by simply counting the number of adults and infants with VOLUME greater than the volume threshold of interest.

Code for calculating the adult harvest proportion using both approaches is provided.

```{r Part_7d}

(adu_max_vol_prop <- (1 - prop.adults)[which.max(smooth.difference)])  # [1] 0.7416332
# OR,
sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] >
      volume.value[which.max(smooth.difference)]) / total.adults # [1] 0.7416332

# Find prop infants
(inf_max_vol_prop <- (1 - prop.infants)[which.max(smooth.difference)])

```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

***####  Section 8:  ####***

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions. You may use either the 'indexing' or 'count' approach, or both.

```{r Part_8a}

zero_a1_cutoff <- volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786
zero_a1_cutoff

# Proportion of infants
inf_a1_prop <- sum(mydata[mydata$TYPE == "I", "VOLUME"] > zero_a1_cutoff) / total.infants
inf_a1_prop

# Proportion of infants
adu_a1_prop <- sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] > zero_a1_cutoff) / total.adults
adu_a1_prop
```

(8)(b) Next, append one (1) more vertical A-B line to our (6)(d) graph. This time, showing the "zero A1 infants" cutoff from (8)(a). This graph should now have five (5) A-B lines:  "protect all infants," "median infant," "median adult," "max difference" and "zero A1 infants."

```{r Part_8b, fig.width = 12}
# Create graph
ggplot() +
  geom_line(mapping = aes(x = volume.value, y = prop.infants, col = "Infants")) +
  geom_line(mapping = aes(x = volume.value, y = prop.adults, col = "Adults")) +
  labs(title = "Proportion of Adults and Infants Protected",
       x = "Volume",
       y = "Proportion",
       col = "") + 
  scale_color_manual(breaks = c("Adults", "Infants"),
                     values = c("darkblue", "darkred")) +
  geom_vline(xintercept = med_inf_vol) +
  annotate("text", label=paste("Med Inf\n", round(med_inf_vol, 1)), x=med_inf_vol+30, y=.45) +
  geom_vline(xintercept = med_adu_vol) +
  annotate("text", label=paste("Median Adl\n", round(med_adu_vol, 1)), x=med_adu_vol+40, y=.45) +
  geom_vline(xintercept = max_inf_vol) +
  annotate("text", label=paste("Protect all infants\n", round(max_inf_vol, 1)), x=max_inf_vol+60, y=.95) +
  geom_vline(xintercept =  volume.value[which.max(smooth.difference)]) +
  annotate("text", label=paste("Max dif\n", round(volume.value[which.max(smooth.difference)], 1)),
             x = volume.value[which.max(smooth.difference)]+30, y = .77) +
  geom_vline(xintercept = zero_a1_cutoff) +
  annotate("text", label=paste("Zero\nA1 inf\n", round(zero_a1_cutoff, 1)), x=zero_a1_cutoff+25, y = .65) +
  geom_hline(yintercept = .50, linetype = 2) +
  theme(legend.position = "inside",
        legend.justification = c(.95, .87),
        legend.background = element_blank(),
        plot.title = element_text(face = "bold", size = 12, hjust = 0)
        )

```


***#### Section 9: ####***

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (6), (7) and (8) on this plot and label each. 

```{r Part_9, fig.width = 8.5}
# Create plot
ggplot() +
  geom_line(aes(x = (1 - prop.infants), y = (1 - prop.adults)), col = "darkblue", linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, col = "darkred", linewidth = 1, linetype = 2) +
  geom_point(aes(x = max_inf_prop, y = max_adu_prop), size = 5, shape = 1, col = "black") +
  annotate("text", label=paste("Protect all infants\n vol =", round(max_inf_vol, 1)), 
           x = max_inf_prop+.09, y = max_adu_prop-.05) +
  geom_point(aes(x = prop_inf_har2, y = prop_adu_har2), size = 5, shape = 1, col = "black") +
  annotate("text", label=paste("Median Adult\n vol =", round(med_adu_vol, 1)), 
           x = prop_inf_har2+.069, y = prop_adu_har2-.051) +
  geom_point(aes(x = inf_max_vol_prop, y = adu_max_vol_prop), size = 5, shape = 1, col = "black") +
  annotate("text", label=paste("Max difference\n vol =", round(volume.value[which.max(smooth.difference)], 1)),
           x = inf_max_vol_prop+.0765, y = adu_max_vol_prop-.051) +
  geom_point(aes(x = inf_a1_prop, y = adu_a1_prop), size = 5, shape = 1, col = "black") +
  annotate("text", label=paste("Zero A1 Infants\n vol =", round(zero_a1_cutoff, 1)),
           x = inf_a1_prop-.065, y = adu_a1_prop+.05) +
  geom_point(aes(x= prop_inf_har1, y = prop_adu_har1), size = 5, shape = 1, col = "black") +
  annotate("text", label=paste("Median Infants\n vol =", round(med_inf_vol, 1)), 
           x = prop_inf_har1+.0765, y = prop_adu_har1-.051)
```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package. Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}
# Calculate AUC
round(auc(x = (1- prop.infants), y = (1 - prop.adults)), 3)
```


***#### Section 10: ####***

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults),
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population

To calculate the total harvest proportions, you can use the 'count' approach, but ignoring TYPE; simply count the number of individuals (i.e. rows) with VOLUME greater than a given threshold and divide by the total number of individuals in our dataset.
 	
```{r Part_10} 	
# Concatenate necessary vars
variables <- c("Protect all infants", 
               "Median Infants", 
               "Median Adults", 
               "Max Differences", 
               "Zero A1 Infants")

# Concatenate volume sizes
volumes <- c(round(max_inf_vol, 1), 
             round(med_inf_vol, 1), 
             round(med_adu_vol, 1), 
             round(volume.value[which.max(smooth.difference)], 1), 
             round(zero_a1_cutoff, 1)
             )

# Concatenate TPR
TPR <- c(round(max_adu_prop, 3),
         round(prop_adu_har1, 3),
         round(prop_adu_har2, 3),
         round(adu_max_vol_prop, 3),
         round(adu_a1_prop, 3)
         )

# Concatenate FPR
FPR <- c(round(max_inf_prop, 3),
         round(prop_inf_har1, 3),
         round(prop_inf_har2, 3),
         round(inf_max_vol_prop, 3),
         round(inf_a1_prop, 3)
         )

# Create totalHavest values
totalhar_max_inf_prop <- round(sum(mydata$VOLUME > max_inf_vol) / (total.adults + total.infants), 3)
totalhar_med_inf_prop <-  round(sum(mydata$VOLUME > med_inf_vol) / (total.adults + total.infants), 3)
totalhar_med_adu_prop <- round(sum(mydata$VOLUME > med_adu_vol) / (total.adults + total.infants), 3)
totalhar_max_diff_prop <- round(sum(mydata$VOLUME > volume.value[which.max(smooth.difference)])/ (total.adults + total.infants), 3)
totalhar_zero_a1_prop <- round(sum(mydata$VOLUME > zero_a1_cutoff) / (total.adults + total.infants), 3)

# Concatenate totalHarvest column
totalhar <- c(totalhar_max_inf_prop,
                  totalhar_med_inf_prop,
                  totalhar_med_adu_prop,
                  totalhar_max_diff_prop,
                  totalhar_zero_a1_prop)

# Table
table <- kable(cbind(variables, volumes, TPR, FPR, totalhar), col.names = c("", "Volume", "TPR", "FPR", "totalHarvest"),
              padding = 8)
table
```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the five cutoffs determined in this assignment.**   

***Answer: When examining the ROC curve, it is evident that several potential "cutoffs" can be selected based on the desired balance between sensitivity and specificity. In this case, the five cutoffs assessed each offer different trade-offs, with some focusing on maximizing the number of infants protected, while others aim to reduce the overall harvest volume by capturing a higher proportion of adult abalone. The most conservative harvesting approach is based on Protecting All Infants (526.64 cm^3), followed by the Median Adults (384.56 cm^3), Max Difference (262.14 cm³), Zero A1 Infants (206.79 cm^3), and, finally, the most aggressive, Median Infants (133.82 cm^3). These cutoff points are simply suggested options for abalone harvesting. The Max Difference (262.14 cm^3) appears to be the most effective method for harvesting both adult and infant abalones without risking the overharvesting of infants (i.e., false positives).***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for  implementation of a cutoff?  
4. What suggestions would you have for planning future abalone studies of this type? 

***Answer: I would outline various harvesting choices and their associated trade-offs for abalones. After presenting these options, I would recommend using the Max Difference method as the optimal approach for abalone harvesting, as it represents the best practice. Based on the volume findings, we can more accurately classify abalone and identify which ones should be harvested for research purposes. However, the study's limitations could arise from factors such as the abalone habitat, which may lead to non-random sampling. It’s also worth noting that infant abalones exhibit more variable characteristics compared to adult abalones. One possible approach to implementing cutoffs is to minimize the harvest of infant abalones, particularly avoiding the A1 class, to allow for reproduction and habitat preservation. Additionally, focusing on harvesting more adult abalones could be considered. As mentioned earlier, I would suggest using the Max Difference cutoff point for the abalone harvesting study. For future abalone studies, I recommend employing random sampling across different abalone habitats, considering environmental effects on abalones. Each habitat should be thoroughly analyzed in terms of abalone classes, type, volume, sex, etc., and these factors should be compared across various habitats.***