---
title: 'Pasaye_Joshua'
output:
  html_document: default
---

```{r setup, include=FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)

library(moments)  # install.packages("moments")

```


##### Section 1: ##################################

##### (1) R has probability functions available for use (Kabacoff, Section 5.2.3). Using one distribution to approximate another is not uncommon.

(1)(a) The Poisson distribution may be used to approximate the binomial distribution if n > 20 and np < 7. Estimate the following binomial probabilities using *dpois()* or *ppois()* with probability p = 0.05, and n = 100. Then, estimate the same probabilities using *dbinom()* or *pbinom()*.  Show the numerical results of your calculations.

(i) The probability of exactly 0 successes.

```{r test1ai}
# Inpput known vaues
p <-  0.05
n <- 100

# ppois binomial probabilities
dpois_0 <- dpois(x = 0, lambda = n*p)

# pbinom binomial probabilities
dbinom_0 <- dbinom(x = 0, size = n, prob = p)

# Print numerical results
sprintf("dpois: %.5f vs dbinom: %.5f", dpois_0, dbinom_0)
```

(ii) The probability of fewer than 7 successes. Please note the following, taken from the Binomial Distribution R Documentation page, regarding the "lower.tail" argument:

lower.tail	logical; if TRUE (default), probabilities are P[X ??? x], otherwise, P[X > x].

```{r test1aii}
# Inpput known vaues
p <-  0.05
n <- 100

# ppois binomial probabilities
ppois_6 <- ppois(q = 6, lambda = n*p)

# pbinom binomial probabilities
pbinom_6 <- pbinom(q = 6, size = n, prob = p)

# Print numerical results
sprintf("ppois: %.5f vs pbinom: %.5f", ppois_6, pbinom_6)
```

The binomial may also be approximated via the normal distribution. Estimate the following binomial probabilities using *dnorm()* or *pnorm()*, this time with probability p = 0.2 and n = 100. Then, calculate the same probabilities using *dbinom()* and *pbinom()*. Use continuity correction. Show the numerical results of your calculations.

(iii) The probability of exactly 25 successes.

```{r test1aiii}
# Input known values
p <- 0.2
n <- 100

# pnorm normal distribution
pnorm_25 <- pnorm(q = 25.5, mean = n * p, sd = sqrt(n * p * (1 - p))) - 
  pnorm(q = 24.5, mean = n * p, sd = sqrt(n * p * (1 - p)))

# dbinom discrete distribution
dbinom_25 <- dbinom(x = 25, size = n, prob = p)

# pbinom discrete distribution
pbinom_25 <- pbinom(q = 25.5, size = n, prob = p) - pbinom(q = 24.5, size = n, prob = p)

# Print numerical results
sprintf("pnorm: %.5f vs dbinom: %.5f & pbinom: %.5f", pnorm_25, dbinom_25, pbinom_25)
```

(iv) The probability of fewer than 25 successes. Please note the following, taken from the Normal Distribution R Documentation page, regarding the "lower.tail" argument:

lower.tail	logical; if TRUE (default), probabilities are P[X ??? x], otherwise, P[X > x].

```{r test1aiv}
# Input known values
p <- 0.2
n <- 100

# pnorm normal distribution
pnorm_24.5 <- pnorm(q = 24.5, mean = n * p, sd = sqrt(n * p * (1 - p)))

# dbinom discrete distribution
dbinom_24 <- sum(dbinom(x = 0:24, size = n, prob = p))

# pbinom discrete distribution
pbinom_24.5 <- pbinom(q = 24.5, size = n, prob = p)

# Print numerical results
sprintf("pnorm: %.5f vs dbinom: %.5f & pbinom: %.5f", pnorm_24.5,  dbinom_24, pbinom_24.5)

```

(1)(b) Generate side-by-side barplots using *par(mfrow = c(1,2))* or *grid.arrange()*. The left barplot will show Poisson probabilties for outcomes ranging from 0 to 10. The right barplot will show binomial probabilities for outcomes ranging from 0 to 10. Use p = 0.1 and n = 100. Title each plot,  present in color and assign names to the bar; i.e. x-axis value labels.

```{r test1b}
# Create plots
par(mfrow = c(1,2))
barplot(dpois(x = 0:10, lambda = 100*.1), main = "Poisson Probabilities",
        xlab = "Outcome", ylab = "Probability", col = "darkorange")
barplot(dbinom(x = 0:10, size = 100, prob = .1), main = "Binomial Probabilities", 
        xlab = "Outcome", ylab = "Probability", col = "darkred")
par(mfrow = c(1,1))
```

(1)(c) For this problem, refer to Sections 5.2 of Business Statistics. A discrete random variable has outcomes:  0, 1, 2, 3, 4, 5, 6.  The corresponding probabilities in sequence with the outcomes are: 0.215, 0.230, 0.240, 0.182, 0.130, 0.003, 0.001.  In other words, the probabilty of obtaining "0" is 0.215.  

(i)  Calculate the expected value and variance for this distribution using the general formula for mean and variance of a discrete distribution. To do this, you will need to use integer values from 0 to 6 as outcomes along with the corresponding probabilities. Round your answer to 1 decimal place. 

```{r test1ci}
# Input known values
outcomes <- c(0:6)
probabilities <- c(.215, .230, .240, .182, .130, .003, .001)

# Find expected value
expec_value <-sum(outcomes*probabilities)
sprintf("The expected value of the distribution is: %.1f", expec_value)

# Find variance
variance <- sum((outcomes - expec_value)^2 * probabilities)
sprintf("The variance of the distribution is: %.1f", variance)
```

(ii) Use the *cumsum()* function and plot the cumulative probabilties versus the corresponding outcomes. Detemine the value of the median for this distribution and show on this plot. Note that there are methods for interpolating a median. However, we can identify an appropriate median from our set of our outcomes - 0 through 6 - that satisfies the definition. Creating a stair-step plot of the cumulative probability as a function of the outcomes may be helpful in identifying it.

```{r test1cii}
# Determining cumulative probabilities
cumprob <- cumsum(probabilities)

# Create plot
plot(x = outcomes, y = cumprob, type = "s") 
abline(h = 0.5, lty = 2, col = "red4") # The median is the outcome at which we 'reach' 0.5; i.e. the median, 50th percentile. 
text(x = 2.01, y = 0.52, labels = "Median = 2", pos = 4)
```

##### Section 2: ##################################

#####  (2) Conditional probabilities appear in many contexts and, in particular, are used by Bayes' Theorem. Correlations are another means for evaluating dependency between variables. The dataset "faithful"" is part of the "datasets" package and may be loaded with the statement *data(faithful)*. It contains 272 observations of 2 variables;  waiting time between eruptions (in minutes) and the duration of the eruption (in minutes) for the Old Faithful geyser in Yellowstone National Park.

(2)(a) Load the "faithful" dataset and present summary statistics and a histogram of waiting times.  Additionally, compute the empirical conditional probability of an eruption less than 3.5 minutes, if the waiting time exceeds 70 minutes.

```{r test2a}
data(faithful, package = "datasets")

# Summary of data
summary(faithful)

# Histogram of wait times
hist(faithful$waiting, main = "Waiting Time (minutes)", xlab = "Time (mintues)", ylab = "Frequency",
     col = "steelblue")

# Conditional probability
wait_over_70 <- subset(faithful, subset = waiting > 70)
erupt_3.5 <- subset(wait_over_70, subset = eruptions < 3.5)
cond_prob <- nrow(erupt_3.5)/nrow(wait_over_70)

# Print result
sprintf("The empirical conditional probability is: %.4f", cond_prob)
```

(i) Identify any observations in "faithful" for which the waiting time exceeds 90 minutes and the eruptions last longer than 5 minutes.  List and show any such observations in a distinct color on a scatterplot of all eruption (vertical axis) and waiting times (horizontal axis). Include a horizontal line at eruption = 5.0, and a vertical line at waiting time = 90.  Add a title and appropriate text. 

```{r test2ai}
# Find observations of waiting longer than 90 mins & eruption longer than 5 mins
df2 <- faithful[which(faithful$eruptions > 5 & faithful$waiting > 90), ]
df2

# Create scatterplot
plot(faithful$waiting, faithful$eruptions, main="Eruption vs Waiting Time (minutes)", 
     xlab="Waiting Time (mins)", ylab="Eruption Duration (mins)", pch=19, col="black", ylim = c(0, 5.1))
points(df2$waiting, df2$eruptions, pch=19, col="green")
abline(h=5, col="red", lty=2)
abline(v=90, col="blue", lty=2)

```

(ii)  What does the plot suggest about the relationship between eruption time and waiting time?

***Answer: Wait time and eruption time have a positive correlation, that is the longer the wait, the longer the eruption lasts. Additionally, most of the data lies before the 60 minute wait time and past the 70 minute wait time with the least number of eruption frequencies occur between the 60 to 70 minute wait time. Lastly, there seems to be only one data point that has a wait time longer than 90 minutes and have an eruption last over 5 minutes.    ***

-----

(2)(b) Past research indicates that the waiting times between consecutive eruptions are not independent.  This problem will check to see if there is evidence of this. Form consecutive pairs of waiting times.  In other words, pair the first and second waiting times, pair the third and fourth waiting times, and so forth.  There are 136 resulting consecutive pairs of waiting times.  Form a data frame with the first column containing the first waiting time in a pair and the second column with the second waiting time in a pair. Plot the pairs with the second member of a pair on the vertical axis and the first member on the horizontal axis.

One way to do this is to pass the vector of waiting times - faithful$waiting - to *matrix()*, specifying 2 columns for our matrix, with values organized by row; i.e. byrow = TRUE.

```{r test2b}
# Create matrix with data
mat_wait <- matrix(faithful$waiting, ncol = 2, byrow = TRUE,
                   dimnames = list(NULL, c("FWT", "SWT")))

# Plot pairs
plot(x = mat_wait[, "FWT"], y = mat_wait[, "SWT"], col="darkred", main = "Paired Waiting Times", 
     xlab="First Waiting Time (minutes)", ylab="Second Waiting Time (minutes)", pch = 19)
```

(2)(c) Test the hypothesis of independence with a two-sided test at the 99% confidence level using the Kendall correlation coefficient. The *cor.test()* function can be used to structure this test and specify the appropriate - Kendall's tau - method.

```{r test2c}
# Correlation test
cor.test(mat_wait[, "FWT"], mat_wait[, "SWT"], altnerative = "two.sided", method = "kendall", conf.level = .99)

```

##### Section 3: ##################################

##### (3)  Performing hypothesis tests using random samples is fundamental to statistical inference. The first part of this problem involves comparing two different diets. Using "ChickWeight" data available in the base R, "datasets" package, we will create a subset of the "ChickWeight" data frame. Specifically, we want to create a data frame that includes only those rows where Time == 21 AND Diet == 1 or 3.

```{r test3}

# load "ChickWeight" dataset
data(ChickWeight, package = "datasets")

# There are multiple ways to approach the subsetting task. The method you choose is up
# to you.
result <- ChickWeight[ChickWeight$Time == 21 & (ChickWeight$Diet == 1 | ChickWeight$Diet == 3), ]
head(result)

# The values in your subsetted data frame should match those below:
# > head(df)
#    weight Time Chick Diet
# 12    205   21     1    1
# 24    215   21     2    1
# 36    202   21     3    1
# 48    157   21     4    1
# 60    223   21     5    1
# 72    157   21     6    1

```

##### The data frame, "result", has chick weights for two diets, identified as diet "1" and "3". Use the data frame, "result," to complete the following item.

(3)(a) Display two side-by-side vertical boxplots using par(mfrow = c(1,2)).  One boxplot would display Diet "1" and the other Diet "3". 

```{r test3a}
# Subset data
diet_1 <- subset(result, Diet == "1")
diet_3 <- subset(result, Diet == "3")

# Create plots
par(mfrow = c(1,2))

boxplot(diet_1$weight, main = "Diet 1", ylab = "Chick Weight (g)", col = "steelblue")
boxplot(diet_3$weight, main = "Diet 3", ylab = "Chick Weight (g)", col = "grey")
par(mfrow = c(1,1))
```

(3)(b) Use the "weight" data for the two diets to test the null hypothesis of equal population mean weights for the two diets. Test at the 95% confidence level with a two-sided t-test. This can be done using *t.test()* in R. Do not assume equal variances. Display the results of *t.test()*.

```{r test3b}
t.test(diet_1$weight, diet_3$weight, alternative = "two.sided", conf.level = 0.95, var.equal = FALSE)

```

##### Working with paired data is another common statistical activity. The "ChickWeight" data will be used to illustrate how the weight gain from day 20 to 21 may be analyzed. This time, we will look only at those individuals on Diet == "3". You will need to add code below creating two (2) vectors. One (1) vector should include all the Time == 20 weights of those individuals on Diet == "3"; a second should include all the Time == 21 weights of those individuals on Diet == "3".

```{r test3paired}

# There are multiple ways to approach the subsetting task. The method you choose is up
# to you.

# Create vector for Time == 20
Time_20 <- subset(ChickWeight$weight, ChickWeight$Time == 20 & ChickWeight$Diet == 3)

# Create vector for Time == 21
Time_21 <- subset(ChickWeight$weight, ChickWeight$Time == 21 & ChickWeight$Diet == 3)


# The first six (6) elements of your Time == 20 vector should match those below:
# [1] 235 291 156 327 361 225

```

(3)(c) Present a scatterplot of the Time == 21 weights as a function of the Time == 20 weights.  Include a diagonal line with zero intercept and slope equal to one. Title and label the variables in this scatterplot.  

```{r test3c}
# Create scatterplot
plot(x = Time_20, y = Time_21, main = "Chick Weight for Day 20 vs Day 21",
     xlab = "Weight at Day 20 (grams)", ylab = "Weight at Day 21 (grams)", col = "blue", pch = 19)
abline(a = 0, b = 1, col = "red")

```

(3)(d) Calculate and present a one-sided, 95% confidence interval for the average weight gain from day 20 to day 21. Write the code for the paired t-test and for determination of the confidence interval endpoints. **Do not use *t.test()**, although you may check your answers using this function. Present the resulting test statistic value, critical value, p-value and confidence interval.

```{r test3d}
# Known values
alpha <- 1 - .95

# Find t statistic
d <- Time_21 - Time_20
n <- length(d)
d_bar <- mean(d)
s_d <- sd(d)

t <- d_bar/(s_d/sqrt(n))
sprintf("The test statistic value is: %.4f", t)

# Find critical value
crit_t <- qt(p = 1 - alpha, df = n - 1)
sprintf("The critical value is: %.5f", crit_t)

# Find p-value
p_val <- pt(q = t, df = n - 1, lower.tail = FALSE)
sprintf("The p-value is: %.6f", p_val)

# Find CI
ci <- d_bar - crit_t*(s_d/sqrt(n))
sprintf("The One-sided Confidence Interval is: %.6f < D < inf", ci)
```

##### Section 4: ##################################

##### (4) Statistical inference depends on using a sampling distribution for a statistic in order to make confidence statements about unknown population parameters. The Central Limit Theorem is used to justify use of the normal distribution as a sampling distribution for statistical inference. Using Nile River flow data from 1871 to 1970, this problem demonstrates sampling distribution convergence to normality. Use the code below to prepare the data.  Refer to this example when completing (4)(c) below.

```{r test4}
data(Nile, package = "datasets")

m <- mean(Nile)
std <- sd(Nile)
x <- seq(from = 400, to = 1400, by = 1)
hist(Nile, freq = FALSE, col = "darkred", xlab = "Flow",
     main = "Histogram of Nile River Flow from 1871 to 1970")
curve(dnorm(x, mean = m, sd = std), col = "blue", lwd = 2, add = TRUE)

```

(4)(a) Using Nile River flow data and the "moments" package, calculate skewness and kurtosis. Present a QQ plot and boxplot of the flow data side-by-side using *qqnorm()*, *qqline()* and *boxplot()*; *par(mfrow = c(1, 2))* may be used to locate the plots side-by-side. Add features to these displays as you choose.

```{r test4a}
# Calculate skewness
skewness(Nile)

# Calcuate kurtosis
kurtosis(Nile)

# Create graphs
par(mfrow = c(1, 2))
qqnorm(Nile, col="blue", main="Nile River QQ Plot")
qqline(Nile, col="blue")
boxplot(Nile, col="green", main="Box Plot of Nile Flow",ylab="Flow")
```

(4)(b) Using *set.seed(456)* and the Nile data, generate 1000 random samples of size n = 16, with replacement. For each sample drawn, calculate and store the sample mean. This can be done with a for-loop and use of the *sample()* function. Label the resulting 1000 mean values as "sample1". **Repeat these steps using *set.seed(789)* - a different "seed" - and samples of size n = 64.** Label these 1000 mean values as "sample2". Compute and present the means, sample standard deviations and sample variances for "sample1" and "sample2" in a table with the first row for "sample1", the second row for "sample2" and the columns labled for each statistic.

```{r test4b}
# Set seed
set.seed(456)

# Create first sample dataframe
sample1<-rep(1:1000,0)

for(i in 1:1000) {
  sample1[i]<-mean(sample(Nile, 16,replace=TRUE))
}

# Set seet
set.seed(789)

sample2<-rep(1:1000,0)

for(i in 1:1000) {
  sample2[i] <- mean(sample(Nile, 64, replace=TRUE))
}

# Calculate means, sds, vars for both samples
means <- c(mean(sample1), mean(sample2))
sds <- c(sd(sample1), sd(sample2))
vars <- c(var(sample1), var(sample2))

# Combine into one table
samples <- cbind(means, sds, vars)
colnames(samples) <- c("Mean", "Std Dev", "Variance")
rownames(samples) <- c("sample1", "sample2")
samples

```

(4)(c) Present side-by-side histograms of "sample1" and "sample2" with the normal density curve superimposed. To prepare comparable histograms, it will be necessary to use "freq = FALSE" and to maintain the same x-axis with "xlim = c(750, 1050)", and the same y-axis with "ylim = c(0, 0.025)." **To superimpose separate density functions, you will need to use the mean and standard deviation for each "sample" - each histogram - separately.** 

```{r test4c}

# Create histograms of "sample1" and "sample2" with normal density curves superimposed
par(mfrow = c(1, 2))
hist(sample1, freq=FALSE, xlim=c(750, 1050), ylim=c(0,0.025), col="darkred")
curve(dnorm(x, mean = mean(sample1), sd = sd(sample1)), col = "blue", lwd = 2, add = TRUE)
hist(sample2, freq=FALSE, xlim=c(750, 1050), ylim=c(0,0.025), col="darkgreen")
curve(dnorm(x, mean = mean(sample2), sd = sd(sample2)), col = "orange", lwd = 2, add = TRUE)

```

-----

##### Section 5: ##################################

##### (5)  This problem deals with contingency table analysis. This is an example of categorical data analysis (see Kabacoff, pp. 145-151). The "warpbreaks" dataset gives the number of warp breaks per loom, where a loom corresponds to a fixed length of yarn.  There are 54 observations on 3 variables: breaks	(numeric, the number of breaks), wool (factor, type of wool: A or B), and tension (factor, low L, medium M and high H).  These data have been studied and used for example elsewhere.  For the purposes of this problem, we will focus on the relationship between breaks and tension using contingency table analysis.

(5)(a) warpbreaks is part of the "datasets" package and may be loaded via *data(warpbreaks)*.  Load "warpbreaks" and present the structure using *str()*. Calculate the median number of breaks for the entire dataset, disregarding "tension" and "wool". Define this median value as "median_breaks". Present a histogram of the number of breaks with the location of the median indicated.

Create a new variable "number" as follows:  for each value of "breaks", classify the number of breaks as either strictly below "median_breaks", or the alternative. Convert the "above"|"below" classifications to a factor, and combine with the dataset warpbreaks.  Present a summary of the augmented dataset using *summary()*.  Present a contingency table of the frequency of breaks using the two variables "wool" and "number".  There should be four cells in this table.

```{r test5a}
data(warpbreaks, package = "datasets")

# Present structure
str(warpbreaks)

# Calculate median number of breaks
median_breaks <- median(warpbreaks$breaks)

# Create histogram
hist(warpbreaks$breaks, main = "Histogram of Loom Wrap Breaks", xlab = "Wrap Breaks per Loom")
text(x=37, y=15, paste("Median =", median_breaks), col="blue")
abline(v=median(warpbreaks$breaks),col="blue", lwd=2)

# Create new variable and add to warpbreaks
number <- factor(ifelse(warpbreaks$breaks < median_breaks, "below", "above"))
warpbreaks <- cbind(warpbreaks, number)

# Summary of new dataset
summary(warpbreaks)

# Create contingency table
cont_table <- table(warpbreaks$wool, warpbreaks$number)
cont_table
```

(5)(b) Using the table constructed in (5)(a), test at the 5% level the null hypothesis of independence using the uncorrected *chisq.test()* (Black, Business Statistics, Section 16.2). Show the results of this test and state your conclusions. 

```{r test5b}
# Chisq test
chisq.test(cont_table, correct = FALSE)

# Conclusion:
# The p-value is .4129 which is greater than .05 so we would fail to reject the null hypothesis.
```


(5)(c) 'Manually' calculate the chi-squared statistic and p-value of the table from (5)(a). The *addmargins()* function can be used to add row and column sums to the table; useful for calculating the expected values for each cell. You should be able to match the chi-squared and p-values from (5)(b). The underlying code for the *chisq.test()* function can be viewed by entering *chisq.test* - without parentheses - in the Console. You are given code below to create the table, add row and column sums and calculate the expected values for the for the first of two (2) rows. You will need to add code to calculate the expected values for the second row and the chi-squared. The *pchisq()* function can be used to return the p-value.

```{r test5c}

tbl <- table(warpbreaks$wool, warpbreaks$number)
mar_tbl <- addmargins(tbl)

e11 <- mar_tbl[3, 1] * mar_tbl[1, 3] / mar_tbl[3, 3]
e12 <- mar_tbl[3, 2] * mar_tbl[1, 3] / mar_tbl[3, 3]

# Manual chi-sqaured statistic
chisq_stat <- ((mar_tbl[1,1]-e11)^2/e11) + ((mar_tbl[1,2]-e12)^2/e12) +
  ((mar_tbl[2,1]-e11)^2/e11) + ((mar_tbl[2,2]-e12)^2/e12)
sprintf("The X-sqaured statistic is: %.5f", chisq_stat)

# Manual p-value
p_value <- pchisq(q = chisq_stat, 1, lower.tail = FALSE)
sprintf("The epxected p-value is: %.4f", p_value)
```

(5)(d) Build a user-defined function, using your code for (5)(c).We want to pass our (5)(a) table to our function and have it return the chi-squared statistic and p-value. You're provided with the 'shell' of a function and will need to add code to calculate the expected values, the chi-squared statistic, the p-value and return (i.e. output) the chi-squared and p-value.

```{r 5d}

chisq_function <- function(x) {
  # Code for calculating the expected values
  mar_tbl <- addmargins(x)
  e11 <- mar_tbl[3, 1] * mar_tbl[1, 3] / mar_tbl[3, 3]
  e12 <- mar_tbl[3, 2] * mar_tbl[1, 3] / mar_tbl[3, 3]

  
  # Code for calculating the chi-squared
  chisq_stat <- ((mar_tbl[1,1]-e11)^2/e11) + ((mar_tbl[1,2]-e12)^2/e12) +
  ((mar_tbl[2,1]-e11)^2/e11) + ((mar_tbl[2,2]-e12)^2/e12)
  
  # Code for calculating the degrees of freedom and p-value
  df <- nrow(x)-1
  p_value <- pchisq(chisq_stat, df, lower.tail = FALSE)
    
  # Code to ouput the chi-squared, degrees of freedom and p-value 
  return(list("chi-squared" = chisq_stat,
              "degrees of freedom" = df,
              "p-value" = p_value))
  
}

chisq_function(tbl)

```
